\thispagestyle{empty}
\begin{footnotesize}
\section*{Executive Editor}

Gonzalo Linares Matás,  St. Hugh's College, University of Oxford (UK). Archaeology and Anthropology.
\section*{Editorial Board}
\begin{multicols}{2}
\begin{itemize}
		\setlength{\itemsep}{0pt}%
		\setlength{\parskip}{0pt}%
		\setlength{\parsep}{0pt}%
\item Jennifer Bates, University of Cambridge (UK). South Asian Archaeology.
\item Samara Broglia de Moura, École Pratique des Hautes Études (France). Central Asian Archaeology.
\item Gesualdo Busacca, Stanford University (USA). Near Eastern Archaeology.
\item Christina Carolus, University of California, Berkeley (USA). Archaeobotany.
\item Ximena Chávez Balderas, Tulane University (USA) and Instituto Nacional de Antropología e Historia (México). Mesoamerican Archaeology.
\item Chelsea Colwell-Pasch, Flinders University (Australia). Maritime Archaeology.
\item Amelia Eichengreen, Columbia University (USA). Classical Archaeology.
\item Lucy Gill, Columbia University (USA). Environmental Archaeology.
\item Linda Gosner, Brown University (USA). Classical Archaeology.
\item Alexandra Guglielmi, University College Dublin (Ireland). Iron Age Europe and Roman Archaeology.
\item Jacqueline Jordaan, Pretoria University (South Africa). African Archaeology.
\item Kate Rose, Harvard University (USA). Egyptian Archaeology.
\item Fabio Saccoccio, Università Roma-La Sapienza (Italy). Prehistoric Archaeology.
\item Celeste Samec, Universidad de Buenos Aires (Argentina). South American Archaeology.
\item Erik Shell, University of California, Los Angeles (USA). Classical Archaeology.
\item Rhiannon Stammers, La Trobe University (Australia). Experimental and South African Archaeology.
\item Nick Stephens, Max-Planck Institute for Evolutionary Anthropology (Germany). Human Evolutionary Studies.
\end{itemize}
\end{multicols}


\section*{Editing Team}
\begin{itemize}
		\setlength{\itemsep}{0pt}%
		\setlength{\parskip}{0pt}%
		\setlength{\parsep}{0pt}%
\item Lukas C. Bossert, Humboldt-Universität zu Berlin (Germany). Classical Archaeology.
\item Devin Lea Ward, University of Cambridge (UK). Biological Anthropology.
\item Celeste Samec, Universidad de Buenos Aires (Argentina). South American Archaeology.
\item Samara Broglia de Moura, École Pratique des Hautes Études (France). Central Asian Archaeology.

\end{itemize}
\section*{Reviewing Team}
\begin{multicols}{2}
\begin{itemize}
		\setlength{\itemsep}{0pt}%
		\setlength{\parskip}{0pt}%
		\setlength{\parsep}{0pt}%
\item Constantina Alexandrou, Trinity College Dublin (Ireland). Cypriot Archaeology.
\item Foteini Athanasiadou, University of Oxford (UK). Greek Archaeology.
\item Sabrina Autenrieth, Leiden University (Netherlands). Prehistoric Europe.
\item Rebecca Biermann, State University of New York, Buffalo (USA). Human Evolutionary Studies, African Archaeology.
\item Lukas C. Bossert, Humboldt-Universität zu Berlin (Germany). Classical Archaeology.
\item Guy Bud, University of Oxford (UK). Numismatics, Colonial and Post-Medieval History.
\item Tiffany Cain, University of Pennsylvannia (USA). Historical Archaeology (Mesoamerica).
\item Brian Clark, Rice University (USA). East African Archaeology.
\item Cherene De Bruyn, Pretoria University (South Africa). South African Archaeology, Physical Anthropology.
\item Antonia Edwards-Freshwater, University of Oxford (UK). Near Eastern Archaeology.
\item Francisco Gomes, University of Lisbon (Portugal). Mediterranean Archaeology.
\item Jordan Graham, University of Oxford (UK). Museum Studies, Indigenous Archaeology.
\item Helen Heath, Otago University (New Zealand). Pacific and South-East Asian Archaeology.
\item Jina Heo, University of Wisconsin-Madison (USA). Korean and East Asian Archaeology.
\item Nayeli Jiménez Cano, Universidad Autónoma de Madrid (Spain). Mayan Archaeology, Zooarchaeology.
\item Lisa Johnson, University of California, Berkeley (USA). Mesoamerican Archaeology.
\item Milosz Klosowski, University of Oxford (UK). Classical Archaeology, Art History Studies.
\item Marisol Lang, University of Oxford (UK). Classical Archaeology.
\item Juan Ignacio Morales Hidalgo, Institut de Paleoecologia Humana i Evolució Social (Spain). Prehistoric Archaeology.
\item Pablo Poveda Arias, Universidad de Salamanca (Spain). Medieval Archaeology.
\item Gabriel Prieto, Yale University (USA). South American Archaeology.
\item Katie Rhodes, University of Chicago (USA). Human Genetic Studies.
\item Hannah Ryan, University of Oxford (UK). Scientific Archaeology - Isotopic Studies.
\item Richard Takkou, University of Oxford (UK). Mediterranean and Mesoamerican Archaeology.
\item John Vandergugten, Simon Fraser University (Canada). Social Zooarchaeology, Historical Archaeology.
\item Katerina Velentza, University of Southampton (UK). Maritime Archaeology. Aegean Archaeology.
\item Devin Lea Ward, University of Cambridge (UK). Biological Anthropology.
\item Crista Wathen, University of Oxford (UK). Scientific Archaeology – Dating.
\item Manuel Will, Eberhard Karls Universität Tübingen (Germany). Human Evolutionary Studies, Prehistoric Archaeology.
\end{itemize}
\end{multicols}
\end{footnotesize}
\thispagestyle{empty}